MR-PFU: A suite of R packages for energy conversion chain analysis

### 2023-08-08 paper.Rmd
Working on EA suggestions.

The most important design decision for the suite of `R` packages
involved data format.
We authors are among those who
developed the Physical Supply-Use Table (PSUT) framework,
a matrix-based approach to describing energy flows 
("products" in PSUT terminology) 
from resource extraction to
processing stages
("industries" in PSUT terminology)
and, ultimately, 
to final demand [@Rocco:2016; @Guevara:2017; @Heun:2018; @Aramendia:2022tv].
We chose the PSUT framework as the data format for the new `R` packages, 
because it succinctly describes an entire ECC
for one country and one year
via a set of six matrices
(the "RUVY" matrices).
See the following table.

@@

Further development followed selection of the PSUT framework.
First, the RUVY matrices carry the challenge that 
different countries and years 
have varying energy carriers (products) and
varying energy conversion machines (industries),
meaning that RUVY matrices 
have differing row names, differing column names, and differing sizes.
To get around this challenge, we created the
[matsbyname][link-matsbyname] package [@Heun-matsbyname:2023]
which enables matrix mathematics 
that respects row and column names, 
inserting **0** vectors when needed.
Second,
wanted to be able to perform _matrix_ mathematics 
as easily as _scalar_ mathematics
in `R` data frames
using the [tidyverse][link-tidyverse] syntax [@Wickham:2019]. 
We developed the [matsindf][link-matsindf] package [@Heun-matsindf:2023]
to enable this functionality.
Finally, manipulating row and column names proved to be a challenge, 
especially for RUVY matrices in [matsindf][link-matsindf] data frames, so 
we developed the [RCLabels][link-rclabels] package [@Heun-RCLabels:2023]
to assist.
The table below summarizes these packages, 
all of which are generally useful and available on 
[CRAN][link-cran].

@@

The steps to create the PSUT matrices for each country and each year
are accomplished by a
[targets][link-targets] [@Landau:2021aa]
computation pipeline
available in the [PFUDatabase][link-pfudatabase]
package.
A unique feature of the
[PFUDatabase][link-pfudatabase]
pipeline is an exemplar system that allows
analyses to proceed when allocation or efficiency data for a country are unavailable.
The calculation pipeline in the [PFUDatabase][link-pfudatabase] packge 
allows allocation and efficiency data
for any country and year to be added at any time to improve the database.

>>>>

The most important design decision for the suite of `R` packages
involved data format.
We authors are among those who
developed the Physical Supply-Use Table (PSUT) framework
[@Rocco:2016; @Guevara:2017; @Heun:2018; @Aramendia:2022tv],
an adaptation of economic Input-Output (IO) analysis [@Miller:2009]
to describe energy flows through society.
The Physical Supply-Use Table (PSUT) framework 
is a matrix-based approach that tracks energy flows 
("products" in PSUT terminology) 
from resource extraction to
processing stages
("industries" in PSUT terminology)
and, ultimately, 
to final demand
in a way that is compatible with the IEA's world extended energy balances 
[@International-Energy-Agency:2023aa].
We chose the PSUT framework as the data format for the new `R` packages, 
because it succinctly describes an entire ECC
for one country and one year
via a set of six matrices
(the "RUVY" matrices).
See the following table.

@@

Further development followed selection of the PSUT framework.
First, the RUVY matrices carry the challenge that 
different countries and years 
have varying energy carriers (products) and
varying energy conversion machines (industries),
meaning that RUVY matrices 
have differing row names, differing column names, and differing sizes.
To get around this challenge, we created the
[matsbyname][link-matsbyname] package [@Heun-matsbyname:2023]
which enables matrix mathematics 
that respects row and column names, 
inserting **0** row or column vectors when needed.
Second,
we wanted to be able to perform _matrix_ mathematics 
as easily as _scalar_ mathematics
in `R` data frames
using the [tidyverse][link-tidyverse] syntax [@Wickham:2019]. 
We developed the [matsindf][link-matsindf] package [@Heun-matsindf:2023]
to enable this functionality.
Finally, manipulating row and column names proved to be a challenge, 
especially for RUVY matrices in [matsindf][link-matsindf] data frames, so 
we developed the [RCLabels][link-rclabels] package [@Heun-RCLabels:2023]
to assist.
The table below summarizes these packages, 
all of which are generally useful and available on 
[CRAN][link-cran].

@@

The steps to create the PSUT matrices for each country and each year
are accomplished by a
[targets][link-targets] [@Landau:2021aa]
computation pipeline
available in the [PFUDatabase][link-pfudatabase]
package.
A unique feature of the
[PFUDatabase][link-pfudatabase]
pipeline is an innovative exemplar system that allows
analyses to proceed when allocation or efficiency data for a country are unavailable.
When allocations or efficiencies are missing,
a string of exemplar countries are queried for similar information.
For example, exemplars for Belgium are
France, Europe, and ultimately, the world, in that order.
The calculation pipeline in the [PFUDatabase][link-pfudatabase] package 
allows allocation and efficiency data
for any country and year to be added at any time to improve the database.
